Historical trends in the contribution of six major causes of death to the black-white

### 2017-04-13 Manuscript/Manuscript_COD_Trends.Rmd
moar

@@

- While we found that many state-level trends followed the same pattern of COD-specific contributions, several states were divergent. For every causes of death except communicable disease, Massachusettes and New York exhibited steep declines to the contribution of each cause to the life expectancy gap, implying that decreasing inequality across a broad range of causes has results in eliminating the life expectancy gap in those states.

- On the other hand, DC underwent an increasing life expectancy gap over time and has the largest life expectancy difference of all the states. This analysis

@@

@@

Lit review

@@

CVD
Why did NY and Mass do so well at decreasing the disparity? Why did Wisconsin and Michigan do poorly?

@@

Policy reflections
-CVD still contributes the most to the currently LE gap. Firebaugh et al (2014) say that policy needs to focus on treatment once CVD is realized, not prevention ("The incidence component is negligible, indicating that blacks and whites are equally likely to succumb to heart disease. Heart disease nevertheless account for one quarter of the gap in 2015 because black victims tends to be younger than white victims. Iterventions should focus on identifying heart disease and equalizing treatment. Same goes for Cancers, non-comm diseases. Contrast with HIV/AIDS, prevention is key (since diff is in "incidence", not age at death with comm disease)"
  - can we see this in our age-cod breakdown?/are these conclusions supported there - something about treatment vs. prevention?

@@

Strengths

Limitations
-does not account for migration. Washington DC - have a lot of migration of high SES white folks coming in, and perhaps also relatively high SES black folks moving out
-see the migration literature sent by Sam
-does not account for hispanic ethinicity - demographic shifts in FL, CA, TX for example of black and white hispanics which will pull LE and disparities
-high level of mexican immigration since 1965

Problems specific to DC
-migration's affect on LE over the course of the time period

>>>>

**Men**

```{r stacked_trends_male, echo=F, fig.height=9, fig.width=11}
plot1 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Male"), aes(x = year, y = COD_cont_yrs_mean)) + 
  geom_area(aes(fill = COD_rev), stat = "identity") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + geom_hline(aes(yintercept = -0.4)) + #the axes
  geom_line(data = subset(BlackWhite_results, sex == "Male"), aes(x = year, y = LE_wbgap_mean)) +
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
  theme_classic(base_size = 10) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (%)") +
  xlab("Year (1969-2013)"))

for(i in 1:400){
  if(plot1$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
    plot1$x$data[[i]]$line$width <- 1
  }
}

plot1

plot2 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & year > 2009), aes(x = year, y = COD_cont_yrs_mean)) + 
  geom_area(aes(fill = COD_rev), stat = "identity") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 2009)) + geom_hline(aes(yintercept = -0.4)) + #the axes
  geom_line(data = subset(BlackWhite_results, sex == "Male" & year > 2009), aes(x = year, y = LE_wbgap_mean)) +
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
  theme_classic(base_size = 10) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (%)") +
  xlab("Year (2009-2013)"))

for(i in 1:400){
  if(plot2$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
    plot2$x$data[[i]]$line$width <- 1
  }
}

plot2
```

**Women**

```{r stacked_trends_female, echo=F, fig.height=9, fig.width=11}
plot3 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Female"), aes(x = year, y = COD_cont_yrs_mean)) + 
                      geom_area(aes(fill = COD_rev), stat = "identity") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + geom_hline(aes(yintercept = -0.4)) + #the axes
  geom_line(data = subset(BlackWhite_results, sex == "Female"), aes(x = year, y = LE_wbgap_mean)) +
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
  theme_classic(base_size = 10) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (%)") +
  xlab("Year (1969-2013)"))

for(i in 1:400){
  if(plot3$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
    plot3$x$data[[i]]$line$width <- 1
  }
}

plot3

plot4 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & year > 2009), aes(x = year, y = COD_cont_yrs_mean)) + 
                      geom_area(aes(fill = COD_rev), stat = "identity") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 2009)) + geom_hline(aes(yintercept = -0.4)) + #the axes
  geom_line(data = subset(BlackWhite_results, sex == "Female" & year > 2009), aes(x = year, y = LE_wbgap_mean)) +
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
  theme_classic(base_size = 10) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (%)") +
  xlab("Year (2009-2013)"))

for(i in 1:400){
  if(plot4$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
    plot4$x$data[[i]]$line$width <- 1
  }
}

plot4
```

@@

- While we found that many state-level trends followed the same pattern of COD-specific contributions, several states diverged in ways that either resulted in the near elimination of the life expectancy gap, or a less favorable result. New York and Massachusettes have virtually eliminated the black-white life expectancy gap for both men and women. While these states showed contribution decreases for all major causes of death except communicable disease, both states began with only a minimal contribution of CVD to the mortality gap.

- How do these results compare with states with large black populations: Georgia, Florida, Texas, and California? In all of these states, deaths during the first year of life are relatively higher in blacks and in most cases, deaths from CVD during mid-life is also higher among blacks. California and Florida black young men are dying more of injuries, and black men and women in Florida are dying more of communicable disease.

- These findings can be contrasted with those of Firebaugh et al (2014). These authors distinguish between inequality caused by a difference in CVD mortality incidence vs. that caused by a difference in age-at-death and find that while blacks and whites are equally likely to die of heart disease, blacks are dying at younger ages. [qu: how can the equally likely to die of Heart disease part be true when we have the age std mort rates showing otherwise? [*See page 27 of firebaugh - highlighted section - in dropbbox papers directory*].

*Notes of Firebaugh 2014 paper:*
-CVD still contributes the most to the current LE gap. Firebaugh et al (2014) say that policy needs to focus on treatment once CVD is realized, not prevention ("The incidence component is negligible, indicating that blacks and whites are equally likely to succumb to heart disease. Heart disease nevertheless account for one quarter of the gap in 2015 because black victims tends to be younger than white victims. Interventions should focus on identifying heart disease and equalizing treatment. Same goes for Cancers, non-comm diseases. Contrast with HIV/AIDS, prevention is key (since diff is in "incidence", not age at death with comm disease)"
[*Does this reasoning make sense to you? I'm not sure I follow why a diff in age at death would imply focusing on treatment rather than prevention -- what if blacks get CVD at younger ages than whites? Doesn't focusing on equalizing treatment assume that blacks and whites have the same timing of onset?*]

@@

We used data over 45 years, across two race/ethnicities, and for all age groups. By examining a wide time window, trends were examined that provided context for the contributions estimated in the most recent year. This allowed us to recognize when the estimated contributions for a particular cause were relatively low *because they started low* vs. low *because they trended that way*. Furthermore, our analysis was centred on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages. This is especially important in the light of recent studies that have recieved criticism for narrowly focusing on a specific age which may present a distorted reality [Case & Deaton, Minton IJE 2016, ]. Alongside the decomposition results and age-standardized mortality rates, the scope of our analysis allows readers to examine differences in state-level inequality from many angles. [Methodologically, using the Bayesian autoregressive model, allowed us to estimate mortality rates in strata that were sparsely populated...*something about methods?*]

Our analysis has two main limitations. Firstly, we did not account for hispanic ethinicity, implying that both black and white populations in our study include some portion of hispanics that is known to differ over time and by state. We decided against excluding hispanics from this study because the indicator for ethnicity has been noted to have varying accuracy across states and during the earlier years of data. Thus, states with demographic shifts due to immigration from Mexico such as Florida, California, and Texas, will have estimates of life expectancy and the contribution that are affected. However, we are somewhat reassured by a previous sensitivity analysis, where we found that accounting for hispanic ethinicity did not have a substantial effect on the results [ref Harper 2012]. Secondly, our analysis did not account for the migration of individuals between states. Thus, states with black or white populations that have shifted over time will be affected. This limits the interpretation of results especially for Washington DC, as it has undergone substantial demographic shifts since X.

Problems specific to DC
-migration's effect on LE over the course of the time period

@@

@@

@@

@@
